Synthesizing Quotient Lenses

### 2018-02-23 papers/QuotientLenses/main.tex
many intro typos fixed

Oftentimes, the same data is stored in a variety different formats. For
example, \bibtex{} and EndNote both store bibliography data; spreadsheet data
can be represented as a CSV file or a TSV file; web APIs can return data as a
JSON or
XML object. When different applications manipulate data in different formats,
synchronizing the applications amounts to synchronizing their respective data
formats. One way to synchronize data formats is to use \emph{lenses}, \emph{i.e.,} programs that represent pairs of functions for translating data from
format A to B (the \emph{get} direction) and back again (the \emph{put}
direction).  These lenses, when defined using domain-specific
languages like Boomerang~\cite{boomerang} or
BiGul~\cite{DBLP:conf/pepm/KoZH16}, typically come with a variety of guarantees,
the \emph{lens laws}, that help ensure put and get functions are effective
inverses of one another.
%comprise such synchronization functions, and the ``lens
%laws'' provide restrictions on how these synchronization functions interoperate%.
%Bidirectional languages like Boomerang~\cite{boomerang} and
%BiGul~\cite{DBLP:conf/pepm/KoZH16} use a single expression to encode all of the
%lens's synchronization functions at the same time.

One way to classify lenses is by the laws they obey.  Amongst the most
restrictive lenses are the \emph{bijective lenses}.  These lenses demand
that get and put functions for a bijection, meaning no information
is lost when one translates from data format to another.  While this
strong guarantee can be useful in some contexts, it is more often the case
that some of the information in one or both of the data formats is superfluous.
For instance, one typically does not care to preserve the amount of
whitespace between characters.  In other circumstances, capitalization
or the order of fields in a record or items in a list may be unimportant.
%
%bijective lenses have
%strong correctness guarantees, they are often
%However, these lens laws are too restrictive for many important transformations.
In particular, these lens laws do not permit transformations that put the data
in canonical forms.
For instance, to return to our bibliography example, in \bibtex{},
an author can be specified as
``\cd{lastname, firstname}'' or as ``\cd{firstname lastname}''---it does not matter which of the
two formats is used.
%While it is important to allow authors
%to be input in both ways, we would like these authors to only be output as
%\cd{firstname lastname}, but the lens laws prevent such functions.
{\em Quotient lenses}~\cite{quotientlenses} loosen many of the restrictions
of bijective lenses by introducing an equivalence relation and considering
bijections modulo that relation.  Such quotient lenses make it possible
to align many more data sources than would otherwise be possible by
ignoring details that a programmer views as inessential.
%each of the data formats is quotiented by some equivalence relation and the lens
%laws are taken modulo equivalence.
%This allows lenses to be simpler and more
%flexible. For example, inessential details like whitespace and orderings of
%fields can be ignored when writing the transformations between the formats.

%However, while plain lenses are widely used in both standalone systems and
%popular libraries, Boomerang~\cite{boomerang} is the only implementation
%currently offering quotient lenses. Even though quotient lenses permit simpler
%transformations between the formats, they merely push complexity elsewhere.
Ordinary bijective lenses are fairly tricky to program, especially for
beginners, as they require programmers to think in two directions at once
and learn a new type system, which comes with some tricky ambiguity constraints.
Unfortunately, quotient lenses do not improve the situation:  To
write a quotient lens in Boomerang, programmers must work with standard lens
combinators as well as a system of \emph{canonizers}.
These canonizers encode two functions, one that
maps data represented in the original external format
to a canonical element, and another, in the opposite direction, that chooses
which elements of the original format correspond to values in the canonical
format. Furthermore, if a user wants to define the domain and range of
a lens explicitly (a programming practice that is encouraged), she must
provide regular expressions defining input and output formats.
Consequently, to define a quotient lens, one must often three different sorts
of terms: (1) regular expressions that define the two data formats
to be processed, (2) canonizers to normalize the data
and choose representatives, and (3) a bijective lens maps
one canonical form into another.  Somewhat frustratingly,
these three terms share a fair amount of structure.
%Moreover, writing the lens between the equivalence classes can be
%challenging due to the rather fiddly typechecking, mainly due to
%``non-ambiguity'' conditions in the lens well-formedness rules.

In this paper, we improve the process of programming quotient lenses
in two distinct, but connected ways. First, we
introduce a new language of {\em Quotient Regular Expressions}, or QREs, which
enable a programmer to simultaneously specify a regular expression as well as
a set of canonical representatives for strings that match the regular
expression. With
QREs, programmers do not need to define the data and canonization functions
separately -- by adding simple annotations to ordinary regular expressions,
canonization functions can be inferred.

@@

The following list summarizes the central contributions of our work.
\begin{enumerate}
  \item We introduce {\em Quotient Regular Expressions} (QREs)
  which provide a compact, convenient notation that expresses an equivalence
  relation on a regular languages as well as a canonical set representatives
  for this equavalence relation (Section~\ref{QRE}).
  \item We define a language of {\em QRE lenses}.  This language is
  a subset of the bijective quotient
  lenses whose types are given by QREs. Our main technical contribution
  is a proof that each QRE lens $q$ defines the same transformation as a
  QRE lens $q'$ that is in normal form. (Section~\ref{QRE-lenses}).
  \item Using this normal form, we reduce the problem of {\em synthesizing}
  QRE lenses to the previously studied problem of synthesizing bijective lenses
  between regular languages (Section~\ref{synth}).
  \item We extend the Boomerang {\em implementation} with QREs
  and QRE lens synthesis.  We demonstrate the utility of our
  implementation by using it to
  synthesize QRE lenses between several real-world data formats from the
  {\tt data.gov} database (Section~\ref{impl}).
\end{enumerate}
Sections~\ref{relwork} and~\ref{concl} discuss related and future work.

>>>>

Oftentimes, the same data is stored in a variety different formats. For
example, \bibtex{} and EndNote both store bibliography data; spreadsheet data
can be represented as a CSV file or a TSV file; web APIs can return data as a
JSON or
XML object. When different applications manipulate the same data in different formats,
synchronizing the applications amounts to synchronizing their respective data
formats. One way to synchronize data formats is to use \emph{lenses}, \emph{i.e.,} programs that represent pairs of functions for translating data from
format A to B (the \emph{get} direction) and back again (the \emph{put}
direction).  These lenses, when defined using domain-specific
languages like Boomerang~\cite{boomerang} or
BiGul~\cite{DBLP:conf/pepm/KoZH16}, typically come with a variety of guarantees,
the \emph{lens laws}, that help ensure put and get functions are effective
inverses of one another.  
%comprise such synchronization functions, and the ``lens
%laws'' provide restrictions on how these synchronization functions interoperate%.
%Bidirectional languages like Boomerang~\cite{boomerang} and
%BiGul~\cite{DBLP:conf/pepm/KoZH16} use a single expression to encode all of the
%lens's synchronization functions at the same time.

One way to classify lenses is by the laws they obey.  Amongst the most
restrictive lenses are the \emph{bijective lenses}.  These lenses demand
that get and put functions form a bijection, meaning no information
is lost when one translates from one data format to another.  While this
strong guarantee can be useful in some contexts, it is more often the case
that some of the information in one or both of the data formats is superfluous.
For instance, one typically does not care to preserve the amount of
whitespace between characters.  In other circumstances, capitalization
or the order of fields in a record or items in a list may be unimportant.
%
%bijective lenses have
%strong correctness guarantees, they are often
%However, these lens laws are too restrictive for many important transformations.
%The bijective lens laws do not permit transformations that ignore this
%unimportant information---.
Many data formats have other ad hoc rules as well.  
Returning to our bibliography example, recall that in \bibtex{},
an author can be specified as
``\cd{lastname, firstname}'' or as ``\cd{firstname lastname}''---it does not matter which of the
two formats is used.
%While it is important to allow authors
%to be input in both ways, we would like these authors to only be output as
%\cd{firstname lastname}, but the lens laws prevent such functions.
%
{\em Quotient lenses}~\cite{quotientlenses} loosen many of the restrictions
of bijective lenses by introducing an equivalence relation and considering
bijections modulo that relation.  Such quotient lenses make it possible
to align many more data sources than would otherwise be possible by
ignoring details that a programmer views as inessential.
%each of the data formats is quotiented by some equivalence relation and the lens
%laws are taken modulo equivalence.
%This allows lenses to be simpler and more
%flexible. For example, inessential details like whitespace and orderings of
%fields can be ignored when writing the transformations between the formats.

%However, while plain lenses are widely used in both standalone systems and
%popular libraries, Boomerang~\cite{boomerang} is the only implementation
%currently offering quotient lenses. Even though quotient lenses permit simpler
%transformations between the formats, they merely push complexity elsewhere.
Ordinary bijective lenses are fairly tricky to program, especially for
beginners, as they require programmers to think in two directions at once
and learn a new type system, which comes with some tricky ambiguity constraints.
Unfortunately, quotient lenses do not improve the situation:  To
write a quotient lens in Boomerang, programmers must work with standard lens
combinators as well as a system of \emph{canonizers}.
These canonizers encode two functions, one that
maps data represented in the original external format
to a canonical element, and another, in the opposite direction, that chooses
which elements of the original format correspond to values in the canonical
format. Furthermore, if a user wants to define the domain and range of
a lens explicitly (a programming practice that is encouraged), she must
provide regular expressions defining input and output formats.
Consequently, to define a quotient lens, one must often
construct three different sorts
of terms: (1) regular expressions that define the two data formats
to be processed, (2) canonizers to normalize the data
and choose representatives, and (3) a bijective lens that maps
one canonical form into another.  Somewhat frustratingly,
these three terms share a fair amount of structure.
%Moreover, writing the lens between the equivalence classes can be
%challenging due to the rather fiddly typechecking, mainly due to
%``non-ambiguity'' conditions in the lens well-formedness rules.

In this paper, we improve the process of programming quotient lenses
in two distinct, but connected ways. First, we
introduce a new language of {\em Quotient Regular Expressions}, or QREs, which
enable a programmer to simultaneously specify a regular expression as well as
a set of canonical representatives for strings that match the regular
expression. With
QREs, programmers do not need to define the data and canonization functions
separately---by adding simple annotations to ordinary regular expressions,
canonization functions can be inferred.

@@

The following list summarizes the central contributions of our work.
\begin{enumerate}
  \item We introduce {\em Quotient Regular Expressions} (QREs),
  a compact, convenient notation for defining (1) a
  regular language modulo some equivalence relation, and (2) a canonizer
  for that relation (Section~\ref{QRE}).
  \item We define a language of {\em QRE lenses}.  This language is
  a subset of the full set of bijective quotient
  lenses whose types are given by QREs. Our main technical contribution
  is a proof that our QREs are just as expressive as the full set of bijective
  quotient lenses.
  % In other words, if there eeach QRE lens $q$ defines the same transformation as a
  %QRE lens $q'$ that is in normal form.
  (Section~\ref{QRE-lenses}).
  \item We reduce the problem of {\em synthesizing}
  QRE lenses to the previously studied problem of synthesizing bijective lenses
  between regular languages (Section~\ref{synth}).
  \item We extend the Boomerang {\em implementation} with QREs
  and QRE lens synthesis.  We demonstrate the utility of our
  implementation by using it to
  synthesize QRE lenses between data formats drawn from the
  {\tt data.gov} database (Section~\ref{impl}).
\end{enumerate}
Sections~\ref{relwork} and~\ref{concl} discuss related and future work.
